---
ztitle: "Association Rules"
author: "Kyriakos Chatzidimitriou (kyrcha@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    highlight: textmate
    includes:
      after_body: footer.html
    css: style.css
  pdf_document:
    includes:
      before_body: head.tex
      after_body: footer.tex
---

## Association Rules

The association rules extraction algorithm is included in the `arules` library. The supermarkt transaction data to be used for executing market basket analysis can be found in the Grocery Shopping datasets page of ACM RecSys. More specifically we will use the Belgium retail market dataset.

```{r}
library(arules)
fileURL <- "http://fimi.ua.ac.be/data/retail.dat.gz"
download.file(fileURL, destfile="retail.data.gz", method="curl")
# Read the data in basket format
trans = read.transactions("retail.data.gz", format = "basket", sep=" ");
inspect(trans[1:10])
summary(trans)
```

After reading succesfully the transaction we go ahead with our analysis:

```{r}
rules <- apriori(trans, parameter = list(support = 0.01, confidence = 0.6))
```

```{r}
quality(rules) <- round(quality(rules), digits=3)
rules
```

84 rules were generated. To see them use `inspect`.

```{r}
inspect(rules)
```

For a lot more information check the  [Introduction to arules](https://mran.revolutionanalytics.com/web/packages/arules/vignettes/arules.pdf) vignette.