---
title: "DBSCAN"
author: "Kyriakos Chatzidimitriou (kyrcha@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    highlight: textmate
    includes:
      after_body: footer.html
    css: style.css
  pdf_document:
    includes:
      before_body: head.tex
      after_body: footer.tex
---

## DBSCAN

DBSCAN is a density based clustering algorithm. To use it we install the library `dbscan` and load it:

```{r}
library(dbscan)
data <- iris[,-5]
clustering = dbscan(data, eps = 0.5, minPts = 5)
clustering
```

The algorithm returns the clsuters and whether there are any noise points. The noise points are denoted with 0. To get the groupings:

```{r}
clustering$cluster
```

The classes and clusters comparison:

```{r}
table(clustering$cluster, iris$Species)
```

Another package that contains dbscan is the `fpc` package.

## An example specialized for DBSCAN

Lets create and plot the dataset

```{r}
x1 <- rnorm(1000,0,3)
y1 <- rnorm(1000,-5,2)
x2 <- runif(1000, -20, 20)
y2 <- -0.05*x2^2 + 10 + runif(1000, -1, 1)
y <- c(y1,y2)
x <- c(x1,x2)
plot(x,y)
```

### k-means vs. dbscan

From theory we know that k-means will not be able to identify the two clusters.

```{r}
demo <- data.frame(x,y)
kmeansClust <- kmeans(demo, 2)
plot(demo, col = kmeansClust$cluster + 1, main="kmeans")
dbscanClust <- dbscan(demo, eps = 1, minPts = 12)
plot(demo, col = dbscanClust$cluster + 1, pch = ifelse(dbscanClust$cluster, 1, 4), main="dbscan")
```


