---
title: "Hierarchical"
author: "Kyriakos Chatzidimitriou (kyrcha@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    highlight: textmate
    includes:
      after_body: footer.html
    css: style.css
  pdf_document:
    includes:
      before_body: head.tex
      after_body: footer.tex
---

## Hierarchical Clustering

For hierarchical clustering we will use the iris dataset. To make the clustering dendrogram less crowded we will select 30 instances from the iris dataset.

```{r}
indices <- sample(1:nrow(iris), 30)
dataset <- iris[indices, -5]
```

To do hierachical clustering we will use the `hclust` method:

```{r}
hc <- hclust(dist(dataset), method="ave")
```

We have different options to use in the `method` argument to define the distance between the clusters as:

  - the distance between the closests points of the clusters (`single`)
  - the distance between the farthests points of the clusters (`complete`)
  - the average distance between all points in the clusters (`average`)
  - the distance between the centroids of the clusters (`centroid`)
  - and others (see `?hclust` for a list)

To plot the dendrogram and cut the tree into three clusters:

```{r}
# Plot the dendrogram
plot(hc, hang = -1, labels=iris$Species[indices])
# cut tree into 3 clusters
rect.hclust(hc, k=3)
# groups contain the distribution of the samples to the clusters/groups
groups <- cutree(hc, k=3)
```

## Using single

```{r}
hc <- hclust(dist(dataset), method="single")
# Plot the dendrogram
plot(hc, hang = -1, labels=iris$Species[indices])
# cut tree into 3 clusters
rect.hclust(hc, k=3)
# groups contain the distribution of the samples to the clusters/groups
groups <- cutree(hc, k=3)
```
