---
title: "k-Nearest Neighbors"
author: "Kyriakos Chatzidimitriou"
date: "November 7, 2016"
output: html_document
#  html_document:
#    highlight: textmate
#    includes:
#      after_body: footer.html
#    css: style.css
#  pdf_document:
#    includes:
#      after_body: footer.tex
---

## Introduction

The goal of this notebook is to introduce the k-Nearest Neighbors instance-based learning model in R using
the class package. For this example we are going to use the [Breast Cancer Wisconsin (Original) Data Set](http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Original%29). The steps for loading and splitting the dataset to training and validation are the same as in the [decision trees notes](dt.html)

```{r include=FALSE}
# download the dataset
fileURL <- "http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.data"
download.file(fileURL, destfile="breast-cancer-wisconsin.data", method="curl")
# read the data
data <- read.table("breast-cancer-wisconsin.data", na.strings = "?", sep=",")
# remove the id column
data <- data[,-1]
# put names in the columns (attributes)
names(data) <- c("ClumpThickness",
"UniformityCellSize",
"UniformityCellShape",
"MarginalAdhesion",
"SingleEpithelialCellSize",
"BareNuclei",
"BlandChromatin",
"NormalNucleoli",
"Mitoses",
"Class")
# make the class a factor
data$Class <- factor(data$Class, levels=c(2,4), labels=c("benign", "malignant"))
# set the seed
set.seed(1234)
# split the dataset
ind <- sample(2, nrow(iris), replace=TRUE, prob=c(0.7, 0.3))
trainData <- data[ind==1,]
validationData <- data[ind==2,]
```

## Training & Prediction

There is a kNN algorithm in the class package.

```{r}
library(class)
```

Because all the attributes are between 1 and 10 there is no need to do normalization between 0 and 1, since no attribute will dominate the others in the distance calculation of kNN. Because kNN accepts the training and testing datasets without the target column, which puts in a 3rd argument, we are going to do some data manipulation to have the data the way the `knn` function likes them (look the manual with ?knn). Also, because no missing values are allowed in kNN, let’s remove those too.

```{r}
trainData <- trainData[complete.cases(trainData),]
validationData <- validationData[complete.cases(validationData),]
trainDataX <- trainData[,-ncol(trainData)]
trainDataY <- trainData$Class
validationDataX <- validationData[,-ncol(trainData)]
validationDataY <- validationData$Class
```

Let’s predict, since there is no need to training when using kNN. The training instances are the model.

```{r}
prediction = knn(trainDataX, validationDataX, trainDataY, k = 1)
```

You can play with the values of k to look for a better model.

## Evaluation

Make the predictions for the validation dataset and print the confusion matrix:

```{r}
cat("Confusion matrix:\n")
xtab = table(prediction, validationData$Class)
print(xtab)
cat("\nEvaluation:\n\n")
accuracy = sum(prediction == validationData$Class)/length(validationData$Class)
precision = xtab[1,1]/sum(xtab[,1])
recall = xtab[1,1]/sum(xtab[1,])
f = 2 * (precision * recall) / (precision + recall)
cat(paste("Accuracy:\t", format(accuracy, digits=2), "\n",sep=" "))
cat(paste("Precision:\t", format(precision, digits=2), "\n",sep=" "))
cat(paste("Recall:\t\t", format(recall, digits=2), "\n",sep=" "))
cat(paste("F-measure:\t", format(f, digits=2), "\n",sep=" "))
```

### Cross Validation

For performing cross-validation we will use the `caret` package. [Here](https://cran.r-project.org/web/packages/caret/vignettes/caret.pdf) you can find a quick guide to caret.



     Define sets of model
     
Write here the main caret stuff

#### Split data in two groups

The function `createDataParitiion` does a stratified random split of the data. Similar to what we did above ourselves (not stratified though). Then we will use the `train` function to build the kNN model.

```{r}
library(caret)
library(mlbench)
data(Sonar)
set.seed(107)
inTrain <- createDataPartition(y = Sonar$Class, p = .75, list = FALSE)
training <- Sonar[ inTrain,]
testing <- Sonar[-inTrain,]
nrow(training)
nrow(testing)
kNNFit <- train(Class ~ ., 
                data = training,
                method = "knn",
                preProc = c("center", "scale"))
print(kNNFit)
```

We can also search for the best _k_ value given the training dataset.

```{r}
kNNFit1 <- train(Class ~ ., 
                data = training,
                method = "knn",
                tuneLength = 15,
                preProc = c("center", "scale"))
print(kNNFit1)
```

To create a 10-fold cross-validation based search of _k_, repeated 3 times we have to use the function `trainControl`:

```{r}
ctrl <- trainControl(method = "repeatedcv", repeats = 3)
kNNFit2 <- train(Class ~ ., 
                data = training,
                method = "knn",
                tuneLength = 15,
                trControl = ctrl,
                preProc = c("center", "scale"))
print(kNNFit2)
plot(kNNFit2)
```

For predictions:

```{r}
knnPredict <- predict(kNNFit2, newdata = testing )
#Get the confusion matrix to see accuracy value and other parameter values
confusionMatrix(knnPredict, testing$Class )
```

Adding more information in the output:

```{r}
ctrl <- trainControl(method = "repeatedcv", repeats = 3, classProbs=TRUE, summaryFunction = twoClassSummary)
kNNFit4 <- train(Class ~ ., 
                data = training,
                method = "knn",
                tuneLength = 15,
                trControl = ctrl,
                preProc = c("center", "scale"))
kNNFit4
plot(kNNFit4)
```

And finally how to use bootstrap with the caret package:

```{r}
knnFit5 <- train(Class ~ ., 
                data = training,
                 method = "knn",
                 preProcess = c("center", "scale"),
                 tuneLength = 10, 
                 trControl = trainControl(method = "boot"))
knnPredict2 <- predict(knnFit5, newdata = testing)
#Get the confusion matrix to see accuracy value and other parameter values
confusionMatrix(knnPredict2, testing$Class)
```

Reference: [http://rstudio-pubs-static.s3.amazonaws.com/16444_caf85a306d564eb490eebdbaf0072df2.html](http://rstudio-pubs-static.s3.amazonaws.com/16444_caf85a306d564eb490eebdbaf0072df2.html)
